### 2022-05-03 newvariant.tex
minor changes

For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution does not change the estimated effectiveness of a constant-strength intervention.
In this same case, however, assuming longer generation intervals decreases the estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals increases the estimated epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
Constant-strength and constant-speed interventions are idealized representations of real-life interventions, which can range from strength-like (e.g., vaccination and social distancing) to speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).

@@

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---this assumption would not apply at the moment of a drastic policy change, but could be applied to the periods immediately before and after.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
We note that $\kappa_x^{-1}$ also corresponds to the shape parameter of a Gamma distribution.
Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution (see \cite{roberts2007model} for detailed discussion).

@@

We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
Inference of the relative strength $\rho = \Rv/\Rw$ from the relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
\end{equation}
Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case, $\rho = \exp(\delta \Gw)$ when $\Gv=\Gw$).

@@

While detailed contact tracing data can help narrow down uncertainties in the generation-interval distributions, there are additional complexities to comparing generation-interval estimates of different variants.
The generation intervals can be measured in two different ways: backward and forward \citep{champredon2015intrinsic}.
The backward distribution starts from the cohort of infectees who were infected at the same time and looks at when their infectors were infected. 
The backward distribution is subject to dynamical biases because we are more likely to observe recent infections (and therefore shorter generation intervals) when the epidemic is growing.
On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
While the forward distribution is more reliable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of the distribution \citep{ali2020serial, nishiura2010time}.
For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.

>>>>

These two idealized interventions, in turn, allow us to understand how holding $\RR$ or $r$ fixed can lead to different conclusions about epidemic control.
For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution does not change the estimated effectiveness of a constant-strength intervention.
In this same case, however, assuming longer generation intervals decreases the estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals increases the estimated epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
Constant-strength and constant-speed interventions are idealized representations of real-life interventions, which can range from strength-like (e.g., vaccination and social distancing) to speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).

@@

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---this assumption would not apply at the moment of a drastic policy change, but could be applied to the periods immediately before and after.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ (equal to the reciprocal of the Gamma shape parameter) \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution (see \cite{roberts2007model} for detailed discussion).

@@

We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
This inference depends on assumptions about the generation-interval distributions of both strains.
Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
\end{equation}
Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case, $\rho = \exp(\delta \Gw)$ when $\Gv=\Gw$).

@@

While detailed contact tracing data can help narrow down uncertainties in the generation-interval distributions, there are additional complexities to comparing generation-interval estimates of different variants.
The generation intervals can be measured in two different ways: backward and forward \citep{champredon2015intrinsic}.
The backward distribution starts from the cohort of infectees who were infected at the same time and looks at when their infectors were infected. 
The backward distribution is subject to dynamical biases because we are more likely to observe recent infections (and therefore shorter generation intervals) when the epidemic is growing.
On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
While the forward distribution is more stable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of either distribution \citep{ali2020serial, nishiura2010time}.
For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.
